---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/tagDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
    - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
output: fost::default_manuscript
---

<!--

Note: keep the story simple and clean, without getting too dense

Outline:

- Primary analysis:
    - Heatmap
    - GEE
        - Include # of both uncorrected and corrected (FDR) significant associations
        - QIC modeling
        - DAG development
        - Look over the entire model (see which variables are significant, etc)
        - Sensitivity analysis with Carb intake? And with adiponectin? (which is
        involved in TAG metabolism)
    - PLS with ISI and HOMA-IR
        - CV for determining number of comps (best is 2)
        - Loadings (of TAGFA) (tho all are the same, use HOMA/ISI. BCF have
        lower corr between measured and predicted) (Use COrr of both components
        (1 and 2))
        - Prediction (of predicted vs real) (in the pls.rmd file, not in
        manuscript (or maybe in ESM?))
        - Use extracted PLS components in GEE (?)
        - Train and test (CV) to see how it predicts in a new dataset (ie. a
        future visit) (done. In pls.rmd file. As ESM?)
- Secondary analysis:
    - PLS-DA (dysgly)
        - Train and test (CV)
        - Loadings (of TAGFA) 
        - Groupings
        - Prediction (misclassification)
    - AROC
        - dysgly with clinical TAG vs the 'best' fatty acid to compare to
        
Possible interpretations:

- muscle IR -> higher TAG
- But longitudinal, since no interaction with time, so maybe the higher TAG is
because of higher IR (since this population is at risk of being IR).
    
-->

# Title page

Title:

Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD

<!-- During time of study. -->
Affiliation:

(1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario, Canada.
(2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
(3) Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
(5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
(6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario, Canada.

Corresponding author:

- Name: Anthony J. Hanley
- Current address:  
    Department of Nutritional Sciences  
    Faculty of Medicine  
    University of Toronto  
    FitzGerald Building, 150 College Street, Room 341  
    Toronto, ON, Canada, M5S 3E2
- Phone number: 416.978.3616
- Fax number:
- Email: anthony.hanley@utoronto.ca

Disclaimers:

Funding support:

# Abstract

# Background

Outline:

- TAG strong predictor of diabetes and CVD risk
- Not well studied how composition plays a role
- Limited longitudinal data

Our objective was to examine the longitudinal role of the specific composition
of the TAG serum fraction on OGTT-derived measures of insulin sensitivity and
beta-cell function.

NEFA may be a mediating factor in association, but that adjustment will not
attenuate this assoc since TAGFA composition is also dictated by other sources
for DNL (e.g. recent meal, DNL from glucose stores, etc.).

# Subjects and Methods

```{r load_data, message=FALSE}
devtools::load_all('.')
ds0yr <- project_data %>% 
    dplyr::filter(VN == 0)
```

```{r caption_setup}
## Include captions below using `captioner` package
fig <- captioner::captioner(prefix = 'Figure')
cite_f <- pryr::partial(fig, display = 'cite')
sfig <- captioner::captioner(prefix = 'Supplemental Figure')
cite_sf <- pryr::partial(sfig, display = 'cite')
tab <- captioner::captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner::captioner(prefix = 'Supplemental Table')
cite_st <- pryr::partial(stab, display = 'cite')
# usage: cite_st('basicChar')

fig_gee <- fig('gee', 'caption here')
fig_consort <- fig('consort', 'caption here')
```

```{r analyze_gee, cache=TRUE}
gee_df <- analyze_gee()
```

```{r analyze_others}
# CV results are shown in pls.Rmd vignette
pls_homa <- analyze_pls(y = 'linvHOMA', cv = FALSE)
pls_isi <- analyze_pls(y = 'lISI', cv = FALSE)

# Correlation between predicted and actual
pls_homa_t <- analyze_pls(y = 'linvHOMA')
pls_isi_t <- analyze_pls(y = 'lISI')

cor_df <- analyze_corr()
```

{{ Use an external source via DOI for reference to these methods? }}

Recruitment for the Prospective Metabolism and Islet Cell Evaluation (PROMISE) 
cohort baseline visit took place between 2004-2006 from London and Toronto, 
Canada. Individuals were selected to participant if they met the eligibility 
criteria of having one or more risk factors for type 2 diabetes mellitus, 
including obesity, hypertension, family history of diabetes, and/or a history of
gestational diabetes or birth of a macrosomic infant. A total of 736 individuals
attended the baseline visit. Subsequent examinations occurred every three years,
with three examination visits completed to date (2004-2006, 2007-2009, and
2010-2013). Participants are contacted annually by telephone. The current study
sample used data on participants who did not have diabetes at baseline, who
returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477; see the CONSORT diagram in `r
cite_sf('consort')`. Metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics are collected at each
examination visit. Research ethics approval was obtained from Mount Sinai
Hospital and the University of Western Ontario, and all participants provided
written informed consent. Data collection methods were standardized across the 2
centres and research nurses were centrally trained.

## Metabolic characterization

Participants completed an 8-12 hour 75g oral glucose tolerance test (OGTT) at 
each examination visit, with blood samples during at fasting, 30 min, and 2 hr 
post-glucose load. Samples were subsequently processed and frozen at -70°C.
Alanine aminotransferase (ALT) was measured using standard laboratory 
procedures. Cholesterol, HDL, and clinically-measured triacylglycerides (TAG) 
were measured using Roche Modular's enzymatic colorimetric tests (Mississauga, 
ON).  Both insulin and glucose were derived from the OGTT at fasting, 30 minute,
and 2 hour time points. Specific insulin was measured with the Elecsys 1010 
(Roche Diagnostics, Basel, Switzerland) immunoassay analyzer and 
electrochemiluminescence immunoassay, which shows 0.05% cross-reactivity to 
intact human pro-insulin and the Des 31,32 circulating split form (Linco Res. 
Inc) and has a coefficient of variation (CV) of 9.3%. Glucose was determined 
using an enzymatic hexokinase (Roche Modular, Roche Diagnostics) with a 
detection range of 0.11 to 41.6 mmol/L and an inter-assay %CV of <1.1% and an
intra-assay %CV of < 1.9%.  All assays were performed at the Banting and Best
Diabetes Centre Core Lab at Mt Sinai Hospital. Impaired fasting glucose (IFG),
impaired glucose tolerance (IGT), and diabetes were categorized using the 2006
WHO criteria [@WHO2006].

TAGFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be 
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of {{ confirm with Richard }} heptadecanoic acid (17:0) was added as an internal
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction---non-esterified fatty acids (NEFA), 
cholesteryl ester, phospholipid, and TAGFA---was isolated using thin layer 
chromatography. Fatty acid methyl esters were separated and quantified using a 
Varian-430 gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a 
Varian Factor Four capillary column and a flame ionization detector. Fatty acid 
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. There
were 22 fatty acids measured in the TAGFA fraction. Given that each lipid
fraction differs in biology and given the complexity of the analyses, findings
for other lipid fractions in this cohort are reported separately (see ref
@Johnston2016a and for the analysis of the phospholipid and cholesteryl ester
fractions and {{next citation}} for the analysis of the NEFA fraction).

## Anthropometrics and sociodemographics

Height, weight, and waist circumference (WC) were measured at all clinic 
examinations using standard procedures.  WC was measured at the natural waist, 
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process. BMI was calculated by dividing weight (kg) by height (m) squared.
Questionnaires administered at each examination determined sociodemographics.
A version of the Modifiable Activity Questionnaire (MAQ) [@Kriska1990]
determined estimated physical activity. The MAQ collects information on leisure
and occupational activity, including intensity, frequency, and duration, over
the past year. Each reported activity from the MAQ was weighted by its
metabolic intensity allowing for the estimation of MET-hours per week.

## Variable calculation and statistical analysis

Insulin sensitivity and beta-cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
reflects whole-body insulin sensitivity [@AbdulGhani2007]. Beta-cell function
was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
the disposition index (but using OGTT values). Each index has been validated
against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].

The primary outcome variables for this analysis were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2, which were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual TAGFA as mole percent (mol%) of the total fraction and as
concentration (nmol/mL). Pearson correlation coefficients were computed to
assess the relationships of individual TAGFA with other continuous variables.

Of note is clinically measured TAG was only quantified on a portion of the
6-year visit (n = ~120 {{ find true number }}).

Generalized estimating equation (GEE) models [@Zeger1986a] were used in the
primary analysis to determine the longitudinal associations between the outcome
variables and the predictor variables. An auto-regressive of order 1 (AR1) 
correlation matrix was chosen for the GEE models as it had the best model fit
and given the longitudinal design, though other matrices (eg. exchangeable) had
similar fit (data not shown). GEE is well suited to longitudinal cohort studies
given it's capacity to handle missed visits. The predictor variables and
continuous covariates were scaled (mean centered and standardized). The TAGFA
variables were classified as *time-independent* (held constant) as they were
measured only at the baseline visit, while the outcome variables and covariates
were set as *time-dependent*. No imputation was conducted on missing values.

Covariates were selected based on previous studies, from directed acyclic graph 
[@Greenland1999a] recommendations, and from quasi-likelihood information criterion 
(QIC) {{citation?}}. `r cite_st('qic')` shows the covariates compared using QIC. While the
final GEE model selected as best fitting differed between insulin sensitivity
and beta-cell function measures, the differences in QIC values were less than
10 between the models, suggesting similar fit. As such, we selected the model
that had the fewest covariates and that had similar fit between the outcome
measures ({{##}}). The final GEE model was adjusted for {{covariates}}. After scaling,
log-transforming, and exponentiating, the GEE estimates are
interpreted as an expected percent difference in the outcome variable for every 
standard deviation (SD) increase in the predictor variable given the covariates 
are held constant. Lastly for the GEE models, we tested for an interaction with
sex, ethnicity, or time by the predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is 
limited in that it cannot analyze the inherent multivariate nature of the 
composition of the TAGFA fraction. To confirm the GEE results in a multivariate 
environment (all TAGFA), partial least squares regression (PLS) {{citation}} was used to
identify the patterns of TAGFA composition against insulin sensitivity and
beta-cell function. For more detailed explanation of GEE and PLS, please see the
{{supplemental}} Methods.

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a] and pls
`r packageVersion('pls')` for PLS {{cite?}}. The R code and extra analyses for
this manuscript is available at {{code doi}}. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].

The number of components to extract from PLS was determined by 10-fold CV.

Dataset was split into a training set and a testing set to determine the
predictability of the given PLS model. Final results presented use the full dataset.a

# Results

```{r inline}
dys <- project_data %>% 
    prep_dys_data() %>% 
    calc_convert_dys()

tagfa <- calc_tagfa_percent()
outcome_chg <- calc_outcome_changes()
female <- calc_discr_npct(ds0yr$Sex, 'Female')
euro <- calc_discr_npct(ds0yr$Ethnicity, 'European')
estci <- calc_gee_estci(gee_df)
fa_cor <- calc_cor()
cor_pls_homa <- calc_pred_corr(pls_homa_t, pls_homa_t$test_data)
cor_pls_isi <- calc_pred_corr(pls_isi_t, pls_isi_t$test_data)
```

## Basic characteristics of the PROMISE cohort

Basic characteristics/demographics

`r tab_basic` shows some basic character

The outcome variables significantly decreased over the 6 years by a median 
`r outcome_chg$chg` (p`r outcome_chg$p`; n=`r outcome_chg$n`).

`r fig_tagfa` shows the distribution of each TAGFA (see `r stab_tagfa` for mean
and SD {{ confirm }}). Three TAGFA contributed 
`r format_rounding(sum(tagfa$pct))`% to the total TAG concentration:
`r tagfa$c[1]`; `r tagfa$c[2]`; and, `r tagfa$c[3]`.

`r fig_corr` shows a heatmap of the correlation of individual TAGFA with the
outcome variables and several basic characteristics.
As expected, nearly all TAGFA had very strong positive correlations (r=
`r fa_cor$rng$TAG`) with clinically-measured TAG. There was also moderate
negative correlations with HDL (r=`r fa_cor$rng$HDL`). For the outcome variables,
the correlations for the insulin sensitivity measures were generally higher 
(1/HOMA-IR: r=`r fa_cor$rng['1/HOMA-IR']`, ISI: r=`r fa_cor$rng$ISI`) than 
for the beta-cell function measures (all r<0.30).

## Generalized estimating equation models

Main effect

There were {{ ## }} significant associations in the unadjusted GEE model shown
in {{ `r sfig_gee` }}. Only a few associations were attenuated after full model
adjustment primarily with the beta-cell function outcomes, as shown in 
`r fig_gee`. As a concentration, nearly all TAGFA had a strong negative
association on 1/HOMA-IR and ISI, and a few had strong negative associations
with IGI/IR and ISSI-2. As a mol%, four TAGFA (14:0, 16:0, 14:1n-7, and 16:1n-7)
had negative associations with 1/HOMA-IR and ISI (between {{ ## }}% lower
insulin sensitivity for every SD increase in the TAGFA), while several more TAGFA had
positive associations with 1/HOMA-IR and ISI (20:0, 18:1n-9, 20:1n-9, 22:1n-9,
18:2n-6, 20:2n-6, 20:4n-6, and 22:5n-3) predicting a {{ ## }}% higher insulin
sensitivity for every SD increase in the TAGFA. 
One TAGFA, 20:2n-6, had a very strong positive association with the
insulin sensitivity measures, with a {{ ## }}% higher insulin sensitivity for
every SD increase in 20:2n-6.


Both clinically-measured TAG and total TAGFA concentration had very strong
negative associations with all outcome variables.

There were no significant interactions by time, sex, or ethnicity for any of the
TAGFA (data not shown).

## Clustering of TAGFA by metabolic measures

Accuracy of predictive ability of PLS components with real measure {{use test maybe?}}.
Correlation of pred vs real in test set.

The predictive ability of the PLS analysis using the TAGFA composition 

`r fig_pls`

The correlation between the predicted values of the outcome variables from the
PLS analysis compared to the actual value was `r cor_pls_homa$r` 
(p`r cor_pls_homa$p`) for 1/HOMA-IR and `r cor_pls_isi$r` (p`r cor_pls_isi$p`) 
for ISI, suggested TAGFA composition has moderate predictive ability for 
insulin sensitivity measures.


`r calc_pred_corr()`

# Discussion

Outline:

- General findings
- Limitations of study
- Previous lit and limits of those
- How this study fills those gaps and new things learned
- Potential clinical application

[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
control than 2hr-OGTT TAG

[@Kotronen2009a] (this wasn't studied, but discussed) higher hepatic DNL may
increase amount of 'shorter' chain FA (16:0, 14:0, etc.) into VLDL and hence
into circulation. So higher of these FA may indicate higher DNL -> problems with
liver function -> maybe hepatic IR -> cyclical?

Discuss my PL and CE paper as well. Maybe also the NEFA one if it is published
by then...

The four FA that cluster together on the PLS also appear to be the same FA that
make up most of DNL, especially in context of higher carb/refined carb in diet
[@Wilke2009a].

[@Lee2015a] 16:1n-7 may be good marker of liver DNL.

# Acknowledgements

## Funding/support

## Conflict of interest

# Tables and figures

```{r table_basic}
table_basic()
```

```{r fig_dist}
plot_tagfa()
```

```{r table_dist}
table_tagfa()
```

```{r fig_heatmap}
plot_heatmap(cor_df)
```

```{r table_corr}
#table_corr(cor_results)
```

```{r table_qic}
table_qic_main()
```

```{r fig_gee}
plot_gee_main(gee_df)
```

```{r table_gee_main}
table_gee_main(gee_df)
```

```{r fig_pls}
gridExtra::grid.arrange(
    plot_pls(pls_homa) + ggplot2::ggtitle('A: 1/HOMA-IR'), 
    plot_pls(pls_isi) + ggplot2::ggtitle('B: ISI'), 
    ncol = 2)
```

# Supplemental Methods

# References
